LCMM, PLS-DA, and LDA exploration

### 2016-02-23 doc/manuscript.Rmd
Added more code/analysis, used functions more, added results + disc re: lcmm+plsda

---
bibliography: 
    - /home/luke/Documents/bibliography/bib/bib-db/nefaDiabetes.bib
    - /home/luke/Documents/bibliography/bib/bib-db/myrefs.bib
csl:
    - /home/luke/Documents/bibliography/styles/american-medical-association-no-et-al.csl
output:
    word_document:
      reference_docx: /home/luke/Templates/journal-templates/ajcn-template.docx
      fig_height: 6 
      fig_width: 5
      fig_caption: true
---

@@

figs/tabs: 
- tab:basic char
- fig:distr
- fig:nefaCorr (sup?)
- fig:varCorr (sup?)
- fig:geeconc%
- fig:lda

supfigs/tabs:
- tab:geeconc%
- tab:distr

TODO:
- Adjust for baseline prediabetes? SenAnalysis?
- Do PLS-LDA instead of LDA, since it is apparently better for this type of
analysis/data

@@

tab_basic <- tab('basicChar', 'Basic characteristics of the PROMISE participants at each of the 3 time points.')
fig_dist <- fig('nefaDist', 'Concentrations (nmol/mL) of each non-esterified fatty acid in PROMISE participants at the baseline visit (2004-2006)')
fig_gee <- fig('gee', 'Longitudinal associations of individual non-esterified fatty acids (mol% and nmol/mL) with insulin sensitivity and beta-cell function using generalized estimating equations over the 6 years in the PROMISE cohort.  Adjusted for {{covariates}}. Outcome variables were log-transformed, predictor variables were scaled, and x-axis values were exponentiated to represent percent difference per SD increase in the fatty acid.  P-values presented were adjusted for the false discovery rate.')
fig_lda <- fig('lda', 'Linear discriminant analysis to identify patterns of NEFA composition between tertiles of log(ISSI-2) in 463 PROMISE participants at the baseline visit (2004-2006).')

sfig_consort <- sfig('consort', 'CONSORT diagram of sample size at each examination visit.')
stab_qic <- stab('qic', 'Comparing generalized estimating equation models adjusting for different covariates using Quasi-Likelihood Information Criterion.')
sfig_corr <- sfig('corr', 'Pearson correlation heatmap of non-esterified fatty acids (nmol/mL) and basic PROMISE participant characteristics for the baseline visit (2004-2006)')
sfig_gee <- sfig('gee', 'Unadjusted generalized estimating equation modeling of the longitudinal association of individual non-esterified fatty acids (mol% and nmol/mL) with insulin sensitivity and beta-cell function over 6 years in the PROMISE cohort. Only adjusted for time. Outcome variables were log-transformed, predictor variables were scaled, and x-axis values were exponentiated to represent percent difference per SD increase in the fatty acid. P-values presented were adjusted for the false discovery rate.')
sfig_lda <- sfig('lda', 'Linear discriminant factors with explained variance against tertiles of log(ISSI-2) in 463 PROMISE participants at the baseline visit (2004-2006).')
```

@@

Despite the fairly extensive literature on NEFA and diabetes, there are
substantial research gaps. Importantly, most of the experimental work using
individual NEFA species has been in animal and cell lines or if in humans has
been acute, infusion studies. For epidemiological studies, few have examined the
role of total, but not individual, NEFA on the pathogenesis of diabetes and
fewer still have been longitudinal, which is key to understanding causal
associations and given the long timescale in diabetes pathophysiology.
Presently, no longitudinal study has been conducted that has examined the role
of the composition of the NEFA fraction on the pathogenesis of diabetes. Our
research objective was to examine the longitudinal role of the NEFA composition
on insulin sensitivity and beta-cell function. Some specific hypotheses included
that higher palmitic acid while lower polyunsaturated fatty acids such as
eicosapentaenoic acid would associate with lower insulin sensitivity and
beta-cell function.

@@

While GEE controls for the longitudinal design of the data, a limitation is that
it can not analyze the inherent multivariate nature of the composition of the
NEFA fraction. As such, after running the GEE models for the longitudinal
analysis, linear discriminant analysis (LDA) was used to identify the patterns
of NEFA composition for tertiles of the metabolic disorders. Since LDA can't
handle longitudinal data, only the baseline data was used and only the outcome
and predictor variables could be included in the analysis. Data from the later
visit dates were used to confirm the results of the LDA analysis.

All analyses were performed using R `r getRversion()` [@Rbase], along with the R
packages geepack `r packageVersion('geepack')` for GEE [@Hoejsgaard2006a] and 
MASS `r packageVersion('MASS')` for LDA. The R code for this manuscript is
available at {{link}}. Results were statistically significance at p<0.05, after
adjusting for multiple testing using the False Discovery Rate [@Benjamini1995a].

@@

```{r varSet}
outcomes_is <- c('linvHOMA', 'lISI')
outcomes_bcf <- c('lIGIIR', 'lISSI2')
outcomes <- c(outcomes_is, outcomes_bcf)
ne_pct <- grep('^pct_ne', names(ds), value = TRUE)
ne_conc <- c('ne_Total', grep('^ne\\d\\d', names(ds), value = TRUE))
covariates_is <- c('VN', 'Waist', 'Sex', 'Ethnicity', 'BaseAge', 'ALT', 'MET', 'AlcoholPerWk')
covariates_bcf <- c('VN', 'Waist', 'Sex', 'Ethnicity', 'BaseAge', 'ALT', 'FamHistDiab')

@@

```{r geeData, cache=TRUE}
gee_results <- bind_rows(
    analyze_gee(gee_df, outcomes_is, ne_pct, covariates_is, 'mol%', adj.p = TRUE),
    analyze_gee(gee_df, outcomes_is, ne_conc, covariates_is, 'nmol/mL', adj.p = TRUE),
    analyze_gee(gee_df, outcomes_bcf, ne_pct, covariates_bcf, 'mol%', adj.p = TRUE),
    analyze_gee(gee_df, outcomes_bcf, ne_conc, covariates_bcf, 'nmol/mL', adj.p = TRUE)
    ) 
gee_results %>% str
# Use substr after this above to make the Xterms go in order.

gee_results_unadj <- bind_rows(
    analyze_gee(gee_df, outcomes, ne_pct, 'VN', 'mol%', adj.p = TRUE),
    analyze_gee(gee_df, outcomes, ne_conc, 'VN', 'nmol/mL', adj.p = TRUE)
    ) %>% 
    mutate(Yterms = factor(Yterms, unique(Yterms)))
```

```{r ldaData}
# include also the 3-yr and 6-yr data for training/testing.
lda_df <- get_lda_data(ds)
lda_results <- lda(BCF ~ ., data = lda_df)
```

```{r inlineResults}
pctne <- calculate_percent_nefa_contribution(ds) 
misclass_npct <- calculate_misclass_npct(lda_results, lda_df)
pctChg.o <- calculate_outcomes_pct_change(ds)

@@

The primary outcome variables declined between `r pctChg.o` over the 6 years.

`r cite_f('nefaDist')` shows the distribution of the NEFA composition. Four 
individual NEFA made up the vast majority (`r sum(pct_nefa$pct)`%) of the total
NEFA fraction. The largest contributor was `r pctne$c[1]`, followed by
`r pctne$c[2]`, `r pctne$c[3]`, and `r pctne$c[4]`.

All individual NEFA as well as the total fraction had weak to no correlations
(r<0.3) with the basic characteristics (see `r cite_sf('corr')`).

@@

Several associations were seen in the unadjusted GEE models, particularly for 
the concentration data. Full model adjustment attenuated most of these 
associations, with total NEFA, 16:0, and 18:2n-6 (all as nmol/mL) had negative 
associations between IGI/IR and ISSI-2 (see `r cite_f('gee')`). The magnitude of
association for each of these variables was fairly consistent for each beta-cell
function measure. For every one SD decrease in any of these three NEFA 
variables, there was a predicted {{magnitude}}% lower IGI/IR and a 
{{magnitude}}% lower ISSI-2 at each clinic visit. None of the NEFA variables as 
a mol% had any association with the outcomes. Adjusting for TG attenuates all 
associations with IGI/IR and ISSI-2 (data not shown), though sample size at the 
6 year visit is restricted to n=128 in those who had TG measured. There were no
significant interaction effects between time and the individual NEFA on any of
the outcome measures (data not shown).

## Discriminant analysis

There was poor discriminatory ability in identifying patterns between tertiles 
of ISSI-2, with only `r misclass_npct` of participants in the baseline visit who
were classified correctly. In general, participants in the 3rd tertile (a high
ISSI-2) had less of all NEFA, though the mean differences between tertiles were
minor (`r cite_f('lda')`). The separation of tertiles of log(ISSI-2) by the two
linear discriminants is shown in `r cite_sf('lda')`. Analyses on the later
visits to confirm the baseline results were not substantially different from the
baseline visit.

@@

Epidemiology [@Johns2014a]

@@

@@

There is substantial experimental evidence highlighting the role of increased 
NEFA and a subsequent increase in insulin resistance via impairment of insulin 
signaling cascades (reviewed in multiple articles 
[@Martins2012a;@Ebbert2013a;@Capurso2012a;@Surampudi2009]). However, in this
longitudinal analysis, we saw no association between any individual or total
NEFA and neither hepatic (1/HOMA-IR) nor whole-body insulin sensitivity (ISI). 
There are some possible explanations for our findings. Previous studies done *in
vivo* were mostly short term infusion 
[@Xiao2009a;@Xiao2006;@Szendroedi2012a;@Kehlenbrink2012a] or pharmocologic
[@Daniele2014a;@Liang2013a] trials and the observational research was generally
cross-sectional [@Miller2012a;@Johns2014a;@Bush2012a]. The limitation of these
studies were the short time periods examined and the experimentally-induced
elevations in NEFA concentrations, making it difficult to assess the causal role
in non-experimental environments over long time frames. Moreover, some studies
have found null or weak associations of NEFA and IR 
[@Gray2013a;@Magkos2012a;@Kehlenbrink2012a], suggesting that either NEFA only 
have a minor role in IR or that NEFA is the effect, but not cause, of the 
underlying IR. Adipose tissue, which release NEFA during fasting, can become 
insulin resistant and thus lipolysis of intracellular TAG by hormone-sensitive 
lipase may not respond to the inhibitory action of insulin. As such, IR may 
develop before elevations in blood NEFA. There is also evidence to suggest that 
insulin may be involved in NEFA update [@Ramos-Roman2012a], further supporting 
the hypothesis that elevated NEFA is a consequence of higher IR.

@@

This is the first study to our knowledge to examine the composition of the NEFA fraction and

@@

Composition of NEFA and diabetes status (x-sec) [@Liu2010a], found 16:0, 18:0,
18:1, 18:2, 18:3 could be useful biomarkers for identifying healthy vs diabetes

Comment on LDA analysis... ...LDA backs up some of the GEE model results,
showing weak discriminatory ability of the NEFA pattern on identifying beta-cell
function... ...While the overall discriminatory ability of the NEFA pattern was
poor, there was a general pattern that most NEFA were lower in individuals with
higher beta-cell function. However, the differences between groups were minor
and confirms the findings from the GEE modeling.

These results confirm the findings from the longitudinal GEE modeling of
individual NEFA variables.

@@

```{r figGEE, dependson='geeData', fig.cap=fig_gee}
gee_results[sort(order(
    gee_results$Xterms)[
        gee_df %>%
    select(matches('ne\\d\\d')) %>%
    gather(FA, Conc) %>%
    transmute(FA = renaming_fa(FA)) %>%
    unique() %>% 
        .[[1]]
    ])]
order(gee_results$Xterms)
plot_gee_results(gee_results)
gee_results %>% filter(Xterms == 'Total')
gee_results %>% str
```

```{r figLDA, fig.cap=fig_lda}
visualize(lda_results, 'means', label.rename.fun = renaming_fa) %>% 
    vision_simple() +
    labs(y = 'Non-esterified fatty acids',
         x = 'Standardized mean of the fatty acid (SD of the nmol/mL)')
```

# Supplemental Material #

@@

Given the number of possible combinations of outcome and predictor variables,
only ISI and ISSI-2 with total non-esterified fatty acids (nmol/mL) were used to
compare various GEE models and to select a final model.  Baseline age was used
as including both original age and the time variable would result in
collinearity.  Column names: QIC is the quasi-likelihood information criteria
(smaller values, eg. larger negative values, indicate a better fit compared to
other models), Delta is the QIC minus the lowest QIC (models with delta <10 are 
considered equivalent).  Models were:

@@

```{r supFigHeatmap, fig.cap=sfig_corr}
ds %>% 
    filter(VN == 0) %>% 
    design('cor', method = 'pearson') %>%
    lay_base(c(outcomes, 'BMI', 'Waist', 'Age', 'lALT', 'lTAG', 'Chol', 'HDL', 'LDL'), ne_conc) %>%
    build() %>%
    polish(rename.vars.fun = function(x) renaming_list(x) %>%
               gsub('^l(ALT|TAG)', 'log(\\1)', .)) %>% 
    trance('heatmap') %>% 
    visualize(colours = c('darkred', 'darkblue'), number.colours = 5) %>% 
    vision_simple() +
    ylab('Non-esterified fatty acids (nmol/mL)')
```

@@

```{r supFigLDA}
visualize(lda_results, orig.data = lda_df, group.name = 'BCF') %>% 
    vision_simple()
```

```{r supTabGEE}
```

>>>>

---
bibliography: 
    - /home/luke/Documents/bibliography/bib/bib-db/nefaDiabetes.bib
    - /home/luke/Documents/bibliography/bib/bib-db/myrefs.bib
csl:
    - /home/luke/Documents/bibliography/styles/vancouver.csl
output:
    word_document:
      reference_docx: /home/luke/Templates/journal-templates/ajcn-template.docx
      fig_height: 6 
      fig_width: 5
      fig_caption: true
---

@@

figs/tabs: 
- tab:basic char
- fig:distr
- fig:nefaCorr (sup?)
- fig:varCorr (sup?)
- fig:geeconc%
- fig:plsda

TODO:
- Bootstrap for the longitudinal table data... like instead of an ANOVA/GEE for
the p-value...?

@@

tab_basic <- tab('basicChar', 'Basic characteristics of the PROMISE participants at each of the 3 time points.')
fig_dist <- fig('nefaDist', 'Concentrations (nmol/mL) of each non-esterified fatty acid in PROMISE participants at the baseline visit (2004-2006)')
fig_gee <- fig('gee', 'Longitudinal associations of individual non-esterified fatty acids (mol% and nmol/mL) with insulin sensitivity and beta-cell function using generalized estimating equations over the 6 years in the PROMISE cohort.  Adjusted for {{covariates}}. Outcome variables were log-transformed, predictor variables were scaled, and x-axis values were exponentiated to represent percent difference per SD increase in the fatty acid.  P-values presented were adjusted for the false discovery rate.')

sfig_consort <- sfig('consort', 'CONSORT diagram of sample size at each examination visit.')
stab_qic <- stab('qic', 'Comparing generalized estimating equation models adjusting for different covariates using Quasi-Likelihood Information Criterion.')
sfig_corr <- sfig('corr', 'Pearson correlation heatmap of non-esterified fatty acids (nmol/mL) and basic PROMISE participant characteristics for the baseline visit (2004-2006)')
sfig_gee <- sfig('gee', 'Unadjusted generalized estimating equation modeling of the longitudinal association of individual non-esterified fatty acids (mol% and nmol/mL) with insulin sensitivity and beta-cell function over 6 years in the PROMISE cohort. Only adjusted for time. Outcome variables were log-transformed, predictor variables were scaled, and x-axis values were exponentiated to represent percent difference per SD increase in the fatty acid. P-values presented were adjusted for the false discovery rate.')
stab_gee <- stab('gee', 'Longitudinal associations of individual non-esterified fatty acids (mol% and nmol/mL) with insulin sensitivity and beta-cell function using generalized estimating equations over the 6 years in the PROMISE cohort.  Adjusted for {{covariates}}. Outcome variables were log-transformed, predictor variables were scaled, and x-axis values were exponentiated to represent percent difference per SD increase in the fatty acid.  P-values presented were adjusted for the BH false discovery rate.')
stab_gee_unadj <- stab('gee_unadj', 'Unadjusted generalized estimating equations models of the longitudinal associations of individual non-esterified fatty acids (mol% and nmol/mL) with insulin sensitivity and beta-cell function over the 6 years in the PROMISE cohort.  Adjusted for time only. Outcome variables were log-transformed, predictor variables were scaled, and x-axis values were exponentiated to represent percent difference per SD increase in the fatty acid.  P-values presented were adjusted for the BH false discovery rate.')
sfig_lcmm <- sfig('lcmm', 'Latent class mixed model analysis to identify individual classes of trajectories for log(ISSI-2) over the 6 years in the PROMISE cohort.')
sfig_plsda_loadings <- sfig('plsda_loadings', 'Pattern loadings from partial least squares discriminant analysis to identify potential clusters of NEFA composition within the classes extracted from the latent class mixed model in 463 participants from the baseline PROMISE visit (2004-2006). The percent explained variance of each component is shown in brackets on each axis.')
sfig_plsda_groups <- sfig('plsda_groups', 'Clustering of extracted components from the partial least squares discriminant analysis on the classes extracted from the latent class mixed model (LCMM) in 463 participants from the baseline PROMISE visit (2004-2006). The percent explained variance of each component is shown in brackets on each axis.')
```

@@

Despite the fairly extensive literature on NEFA and diabetes, there are
substantial research gaps. Importantly, most of the experimental work using
individual NEFA species has been in animal and cell lines or if in humans has
been acute, infusion studies. For epidemiological studies, few have examined the
role of total, but not individual, NEFA on the pathogenesis of diabetes and
fewer still have been longitudinal, which is key to understanding causal
associations and given the long timescale in diabetes pathophysiology.
Presently, no longitudinal study has been conducted that has examined the role
of the composition of the NEFA fraction on the pathogenesis of diabetes. Our
research objective was to examine the role of the NEFA composition
on insulin sensitivity and beta-cell function using longitudinal and clustering
techniques. Some specific hypotheses included that higher palmitic acid while
lower polyunsaturated fatty acids such as eicosapentaenoic acid would associate
with lower insulin sensitivity and beta-cell function.

@@

While GEE controls for the longitudinal design of the data, a limitation is that
it can not analyze the inherent multivariate nature of the composition of the 
NEFA fraction. As such, after running the GEE models for the longitudinal 
analysis, partial least squares discriminant analysis (PLS-DA) was used to
identify the patterns of NEFA composition against beta-cell function (using ISSI-2,
which has most of the associations in the GEE models). Since PLS-DA can't handle
longitudinal data, the longitudinal data was maximized by extracting latent
trajectories of beta-cell function over the 6 years using latent class mixed
models (LCMM). A limitation of these methods is that they can not be adjusted
for covariates.

All analyses were performed using R `r getRversion()` [@Rbase], along with the R
packages geepack `r packageVersion('geepack')` for GEE [@Hoejsgaard2006a], caret
`r packageVersion('caret')` for PLS-DA, and lcmm `r packageVersion('lcmm')` for
LCMM. The R code for this manuscript is available at {{link}}. Results were
statistically significance at p<0.05, after adjusting for multiple testing using
the False Discovery Rate [@Benjamini1995a].

@@

```{r variableSet}
outcomes_is <- c('linvHOMA', 'lISI')
outcomes_bcf <- c('lIGIIR', 'lISSI2')
outcomes <- c(outcomes_is, outcomes_bcf)
ne_pct <- grep('^pct_ne', names(ds), value = TRUE)
ne_conc <- c('ne_Total', grep('^ne\\d\\d', names(ds), value = TRUE))
covariates_is <- c('VN', 'Waist', 'Sex', 'Ethnicity', 'BaseAge', 'ALT', 'MET', 'AlcoholPerWk')
covariates_bcf <- c('VN', 'Waist', 'Sex', 'Ethnicity', 'BaseAge', 'ALT', 'FamHistDiab')

@@

```{r geeData, cache=TRUE}
gee_results <- bind_rows(
    analyze_gee(gee_df, outcomes_is, ne_pct, covariates_is, 'mol%', adj.p = TRUE),
    analyze_gee(gee_df, outcomes_is, ne_conc, covariates_is, 'nmol/mL', adj.p = TRUE),
    analyze_gee(gee_df, outcomes_bcf, ne_pct, covariates_bcf, 'mol%', adj.p = TRUE),
    analyze_gee(gee_df, outcomes_bcf, ne_conc, covariates_bcf, 'nmol/mL', adj.p = TRUE)
    )

gee_results_unadj <- bind_rows(
    analyze_gee(gee_df, outcomes, ne_pct, 'VN', 'mol%', adj.p = TRUE),
    analyze_gee(gee_df, outcomes, ne_conc, 'VN', 'nmol/mL', adj.p = TRUE)
    )
```

```{r lcmmPls}
# include also the 3-yr and 6-yr data for training/testing.
lcmm_results <- analyze_lcmm(ds)
lcmm_data <- combine_lcmm_data(ds, lcmm_results)
plsda_results <- analyze_plsda(lcmm_data)
```

```{r inlineResults}
pctne <- calculate_percent_nefa_contribution(ds) 
misclass_npct <- calculate_plsda_misclass(plsda_results)
pctChg_o <- calculate_outcomes_pct_change(ds)
lcmm_grps <- calculate_ngroups_lcmm(lcmm_results)
```

@@

The primary outcome variables declined between `r pctChg_o` over the 6 years.

`r cite_f('nefaDist')` shows the distribution of the NEFA composition. Four 
individual NEFA made up the vast majority (`r sum(pct_nefa$pct)`%) of the total
NEFA fraction. The largest contributors were `r pctne$c[1]`, `r pctne$c[2]`, 
`r pctne$c[3]`, and `r pctne$c[4]`. All individual NEFA as well as the total
fraction had weak to no correlations (r<0.3) with the basic characteristics (see
`r cite_sf('corr')`).

@@

Several associations were seen in the unadjusted GEE models 
(`r cite_sf('gee')`), particularly for the concentration data. Full model
adjustment attenuated most of these associations, with total NEFA, 16:0, and
18:2n-6 (all as nmol/mL) had negative associations between IGI/IR and ISSI-2
(see `r cite_f('gee')`). The magnitude of association for each of these
variables was fairly consistent for each beta-cell function measure. For every
one SD decrease in any of these three NEFA variables, there was a predicted
{{magnitude}}% lower IGI/IR and a {{magnitude}}% lower ISSI-2 at each clinic
visit. None of the NEFA variables as a mol% had any association with the
outcomes. Adjusting for TG attenuates all associations with IGI/IR and ISSI-2
(data not shown), though sample size at the 6 year visit is restricted to n=128
in those who had TG measured. There were no significant interaction effects
between time and the individual NEFA on any of the outcome measures (data not
shown). Raw values from the GEE models are shown in `r cite_st('gee_unadj')` for
unadjusted models and `r cite_st('gee')` for fully adjusted models.

## Latent trajectory and NEFA pattern recognition

Three latent classes were extracted from LCMM, with `r lcmm_grps[2,2]` 
participants in the high, `r lcmm_grps[1,2]` in the middle, and `lcmm_grps[3,2]`
in the low beta-cell function groups. The trajectories of the three groups are
shown in `r cite_sf('lcmm')`. These groups were used as the response variables
in the PLS-DA analysis. The PLS-DA results show a very poor discriminatory
abililty of the NEFA composition in classifying participants into the correct
beta-cell function latent class. Only `r misclass_npct[2,1]` 
(`r misclass_npct[2,4]`) participants were correctly classified, primarily into the
middle group. The specific NEFA composition was not able to accurately classify
participants who had low or high beta-cell function.
`r cite_sf('plsda_loadings')` shows the loadings of individual NEFA by the
extracted component, while `r cite_sf('plsda_groups')` shows the clustering of the components by ISSI-2 latent class.

@@

In a cross-sectional analysis of the RISCK cohort, total NEFA had a negative
association with insulin sensitivity and, particularly strong with beta-cell
function [@Johns2014a]. The authors further analysis showed that greater
adiposity was not a mediator between the total NEFA and metabolic function
association.

@@

Using cluster techniques, we found no evidence to suggest that any individual
NEFA contributed uniquely to beta-cell function, confirming the results we saw
in the GEE models. This is particularly seen with the poor predictive ability of
the NEFA composition to classify between groups of beta-cell function.

@@

There is substantial experimental evidence highlighting the role of increased 
NEFA and a subsequent increase in insulin resistance via impairment of insulin 
signaling cascades (reviewed in multiple articles 
[@Martins2012a;@Ebbert2013a;@Capurso2012a;@Surampudi2009]). However, in this
longitudinal analysis, we saw no association between any individual or total
NEFA and neither hepatic (1/HOMA-IR) nor whole-body insulin sensitivity (ISI). 
There are some possible explanations for our findings. Previous studies done *in
vivo* were mostly short term infusion 
[@Xiao2009a;@Xiao2006;@Szendroedi2012a;@Kehlenbrink2012a] or pharmocologic
[@Daniele2014a;@Liang2013a] trials and the observational research was generally
cross-sectional [@Miller2012a;@Johns2014a;@Bush2012a]. The limitation of these
studies were the short time periods examined and the experimentally-induced
elevations in NEFA concentrations, making it difficult to assess the causal role
in non-experimental environments over long time frames. Moreover, some studies
have found null or weak associations of NEFA and IR 
[@Gray2013a;@Magkos2012a;@Kehlenbrink2012a; @Johns2014a], suggesting that either
NEFA only have a minor role in IR or that NEFA is the effect, but not cause, of
the underlying IR. Adipose tissue, which release NEFA during fasting, can become
insulin resistant and thus lipolysis of intracellular TAG by hormone-sensitive 
lipase may not respond to the inhibitory action of insulin. As such, IR may 
develop before elevations in blood NEFA. There is also evidence to suggest that 
insulin may be involved in NEFA update [@Ramos-Roman2012a], further supporting 
the hypothesis that elevated NEFA is a consequence of higher IR.

@@

This is the first study to our knowledge to examine the composition of the NEFA
fraction and longitudinal measures of metabolic functioning.

@@

Using complex measurement and statistical techniques that were rigorously
applied {{? reword}} to the composition of the NEFA fraction, Liu *et al.*
-[@Liu2010a] found that 16:0, 18:0, 18:1, 18:2, 18:3 may be useful biomarkers
for identifying healthy compared to diabetic individuals. However, a limitation
of this study was the smaller sample size {{confirm}} and the cross-sectional
design.

@@

```{r figGEE, dependson='geeData', fig.cap=fig_gee}
plot_gee_results(gee_results)
```

# Supplemental Material

@@

Given the number of possible combinations of outcome and predictor variables,
only ISI and ISSI-2 with total non-esterified fatty acids (nmol/mL) were used to
compare various GEE models and to select a final model.  Baseline age was used
as including both the original age and the time variable would result in
collinearity.  Column names: QIC is the quasi-likelihood information criteria
(smaller values, eg. larger negative values, indicate a better fit compared to
other models), Delta is the QIC minus the lowest QIC (models with delta <10 are 
considered equivalent).  Models were:

@@

```{r supFigHeatmap, fig.cap=sfig_corr}
plot_heatmap(ds)
```

@@

```{r supFigLDA}

```{r supTabGEE}
table_gee(gee_results, stab_gee)
table_gee(gee_results_unadj, stab_gee_unadj)
```

```{r supFigLCMM, fig.cap=}
plot_lcmm_results(lcmm_data)
```

```{r supFigPlsLoadings, fig.cap=}
plot_plsda_loadings(plsda_results)
```

```{r supFigPlsGrp}
plot_plsda_grouping(plsda_results)
```
